---
title: "615 Midterm"
author: "Amie Thomas"
date: "2023-10-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("sf")
library(readr)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(sf)
library(leaflet)
```

###<NOAA>###


#load in the NOAA data and merge
```{r}
Mydata <- read_csv("StormEvents_details-ftp_v1.0_d2020_c20230927.csv.gz")
Mydata_2 <- read_csv("StormEvents_details-ftp_v1.0_d2021_c20231017.csv.gz")
noaa_20_21 <- rbind(Mydata, Mydata_2)
#how much damage was done
```

#get rid of columns not of interest
```{r}
noaa_20_21 <- Mydata |>
  filter(EVENT_TYPE == "Flood") |>
   select(-MAGNITUDE, -MAGNITUDE_TYPE, -CATEGORY, -TOR_F_SCALE, -TOR_LENGTH, -TOR_WIDTH, 
          -TOR_OTHER_WFO, -TOR_OTHER_CZ_STATE, -TOR_OTHER_CZ_FIPS, -TOR_OTHER_CZ_NAME, -BEGIN_RANGE,
          -BEGIN_AZIMUTH, -END_RANGE, -END_AZIMUTH, -CZ_TIMEZONE, -DATA_SOURCE, -WFO)
```

#change damage data into useable intergers
```{r}
noaa_20_21$DAMAGE_PROPERTY <- str_replace_all(noaa_20_21$DAMAGE_PROPERTY, "[^0-9.]", "")
noaa_20_21$DAMAGE_PROPERTY <- as.numeric(noaa_20_21$DAMAGE_PROPERTY)
noaa_20_21$DAMAGE_PROPERTY <- round(noaa_20_21$DAMAGE_PROPERTY)
options(scipen = 999)
noaa_20_21$DAMAGE_PROPERTY <- noaa_20_21$DAMAGE_PROPERTY*100000


noaa_20_21$DAMAGE_CROPS <- str_replace_all(noaa_20_21$DAMAGE_CROPS, "[^0-9.]", "")
noaa_20_21$DAMAGE_CROPS <- as.numeric(noaa_20_21$DAMAGE_CROPS)
noaa_20_21$DAMAGE_CROPS <- round(noaa_20_21$DAMAGE_CROPS)
options(scipen = 999)
noaa_20_21$DAMAGE_CROPS <- noaa_20_21$DAMAGE_CROPS*100000

```

#filter out events with no damage
#find areas with the most damage to property
```{r}
noaa_20_21 <- arrange(noaa_20_21, desc(DAMAGE_PROPERTY))
noaa_20_21 <- noaa_20_21 %>%
  filter(DAMAGE_PROPERTY != 0 | DAMAGE_CROPS != 0)
```

#Questions
what types of flood causes result in the most damage
what months have the most floods
what types of flood causes happen in which months
what states have the most damage
what states have what types of flood causes
can I look at the spread of the event

###<FEMA>###

#load in the FEMA data and merge
```{r}
Mydata_3 <- read_csv("FemaWebDisasterSummaries.csv")
Mydata_4 <- read_csv("DisasterDeclarationsSummaries.csv")
fema_data <- left_join(Mydata_3, Mydata_4, "disasterNumber")
#how much relief was sent
```

#filter to keep only columns of interest
```{r}
fema_data$incidentBeginDate <- year(fema_data$incidentBeginDate)
fema_data_int <- fema_data |> 
  filter(incidentType == "Flood") |>
  filter(incidentBeginDate == "2020" | incidentBeginDate ==  "2021")
```



