---
title: "615 Midterm"
author: "Amie Thomas"
date: "2023-10-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(plotly)
library(knitr)
```

###<NOAA>###

#load in the NOAA data and merge
```{r}
Mydata <- read_csv("StormEvents_details-ftp_v1.0_d2020_c20230927.csv.gz")
Mydata_2 <- read_csv("StormEvents_details-ftp_v1.0_d2021_c20231017.csv.gz")
noaa_20_21 <- rbind(Mydata, Mydata_2)
#how much damage was done
```

#get rid of columns and rows not of interest
```{r}
noaa_20_21 <- noaa_20_21 |>
  filter(grepl("Flood", EVENT_TYPE, ignore.case = TRUE))|>
  select(-MAGNITUDE, -MAGNITUDE_TYPE, -CATEGORY, -TOR_F_SCALE, -TOR_LENGTH, -TOR_WIDTH, 
           -TOR_OTHER_WFO, -TOR_OTHER_CZ_STATE, -TOR_OTHER_CZ_FIPS, -TOR_OTHER_CZ_NAME,
           -BEGIN_RANGE,-BEGIN_AZIMUTH, -END_RANGE, -END_AZIMUTH, -CZ_TIMEZONE, -DATA_SOURCE,
           -WFO)

```

#change damage data into useable intergers
```{r}
noaa_20_21$DAMAGE_PROPERTY <- str_replace_all(noaa_20_21$DAMAGE_PROPERTY, "[^0-9.]", "")
noaa_20_21$DAMAGE_PROPERTY <- as.numeric(noaa_20_21$DAMAGE_PROPERTY)
noaa_20_21$DAMAGE_PROPERTY <- noaa_20_21$DAMAGE_PROPERTY*100000


noaa_20_21$DAMAGE_CROPS <- str_replace_all(noaa_20_21$DAMAGE_CROPS, "[^0-9.]", "")
noaa_20_21$DAMAGE_CROPS <- as.numeric(noaa_20_21$DAMAGE_CROPS)
noaa_20_21$DAMAGE_CROPS <- noaa_20_21$DAMAGE_CROPS*100000

```

#filter out events with no damage
```{r}
noaa_20_21 <- noaa_20_21 %>%
  filter(DAMAGE_PROPERTY != 0 | DAMAGE_CROPS != 0)

noaa_20_21 <- noaa_20_21 |>
  mutate(DAMAGE_TOTAL = DAMAGE_CROPS + DAMAGE_PROPERTY)|> #create total damage column
  relocate(DAMAGE_TOTAL, .before = SOURCE)
```

#create death and injury total columns
```{r}
noaa_20_21 <- noaa_20_21 |>
  mutate(DEATH_TOTAL = DEATHS_DIRECT + DEATHS_INDIRECT)|> 
  relocate(DEATH_TOTAL, .before = DAMAGE_PROPERTY)

noaa_20_21 <- noaa_20_21 |>
  mutate(INJURIES_TOTAL = INJURIES_DIRECT + INJURIES_INDIRECT)|> 
  relocate(INJURIES_TOTAL, .before = DEATHS_DIRECT)

noaa_20_21 <- noaa_20_21 |>
  select(-DEATHS_DIRECT, -DEATHS_INDIRECT, -INJURIES_DIRECT, -INJURIES_INDIRECT)
```


#NOAA QUESTIONS

what months have the most floods
```{r}
flood_month <- noaa_20_21 |>
  group_by(MONTH_NAME) |>
  summarise(count = n()) |>
  arrange(desc(count))

custom_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")


flood_month$MONTH_NAME <- factor(flood_month$MONTH_NAME, levels = custom_order)


ggplot(flood_month, aes(x = MONTH_NAME, y = count, fill = MONTH_NAME)) +
  geom_bar(stat = "identity") +
  labs(x = "Month", y = "Count of Floods") +
  ggtitle("Months with most Floods") +
  labs(subtitle = "Years 2020-2021")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

  
```

-What states have what type of floods
```{r}
state_flood_type <- noaa_20_21 |>
  group_by(STATE, EVENT_TYPE)|>
  summarise(count = n())

#organize to get a better idea 
org_sft <- state_flood_type |>
  pivot_wider(names_from = EVENT_TYPE, values_from = count)
  
 ggplot(state_flood_type, aes(x = STATE, y = count, fill = EVENT_TYPE)) +
  geom_bar(stat = "identity") +
  labs(title = "Flood Events by State", x = "State", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))




```

-what types of floods are the most damaging 
```{r}
dam_flood <- noaa_20_21 |>
  group_by(EVENT_TYPE)|>
  summarise(DAMAGE_TOTAL = sum(DAMAGE_TOTAL))

ggplot(data = dam_flood, aes(x = EVENT_TYPE, y = DAMAGE_TOTAL, fill = EVENT_TYPE)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Flood Damage by Event Type (2020-2021)",
    x = "Event Type",
    y = "Total Damage"
  ) +
  theme_minimal()


#How do coastal floods have no damage
```
#I thought that flash floods would do more damage

-deadliest floods
```{r}
dead_flood <- noaa_20_21|>
  group_by(EVENT_TYPE)|>
  summarise(DEATH_TOTAL = sum(DEATH_TOTAL))

ggplot(dead_flood, aes(x = EVENT_TYPE, y = DEATH_TOTAL, fill = EVENT_TYPE)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths by Flood Event Type", x = "Event Type", y = "Total Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  
```

-Causes of the types of floods
```{r}
flood_cause <- noaa_20_21 |>
  group_by(EVENT_TYPE, FLOOD_CAUSE) |>
  summarise(count = n())
 
 ggplot(data = flood_cause, aes(x = EVENT_TYPE, y = FLOOD_CAUSE, fill = log(count + 1))) + #log for visual effect
   geom_tile() +
   scale_fill_gradient(low = "pink", high = "red") +
   labs(
     title = "Cause of Event Type",
     x = "Event Type",
     y = "Flood Cause",
     fill = "Count"
   ) +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

#what states have the most total damage
```{r}
state_dam <- noaa_20_21 |>
  group_by(STATE)|>
  summarise(DAMAGE_TOTAL = sum(DAMAGE_TOTAL))|>
  arrange(desc(DAMAGE_TOTAL))|>
  head(10)

ggplot(data = state_dam, aes(x = reorder(STATE, -DAMAGE_TOTAL), y = DAMAGE_TOTAL, fill = STATE)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 States with the Most Damage (2020-2021)",
    x = "State",
    y = "Total Damage"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

-what types of floods does new york experience 
```{r}
ny_ftype <- noaa_20_21 |>
  group_by(EVENT_TYPE, STATE)|>
  filter(STATE == "NEW YORK")|>
  summarise(count = n())

ggplot(data = ny_ftype, aes(x = EVENT_TYPE, y = count, fill = EVENT_TYPE)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Flood Event Types in New York (2020-2021)",
    x = "Event Type",
    y = "Count"
  ) +
  theme_minimal() 
```

-how many deaths from floods in new york
```{r}
ny_death <- noaa_20_21 |> 
  filter(STATE == "NEW YORK") |>
  summarise(DEATH_TOTAL = sum(DEATH_TOTAL))
#Answer = 1
#no need for visual. Just curious
  #has the most damage but only has one death. what state has the most deaths
```

-deaths from floods by state
```{r}
top_death_state <- noaa_20_21|>
  group_by(STATE)|>
  summarise(DEATH_TOTAL = sum(DEATH_TOTAL))|>
  arrange(desc(DEATH_TOTAL))|>
  head(10)
#what events caused so many deaths in the year 2020-2021 in Tennesee 

ggplot(top_death_state, aes(x = STATE, y = DEATH_TOTAL, fill = STATE)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths by State", x = "State", y = "Total Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for better readability
```

```{r}
#how many floods did Tennesse have
tenn_flood_n <- noaa_20_21 |>
  filter(STATE == "TENNESSEE")|>
  count(EVENT_TYPE)

ggplot(tenn_flood_n, aes(x = EVENT_TYPE, y = n, fill = EVENT_TYPE)) +
  geom_bar(stat = "identity") +
  labs(title = "Flood Event Types in Tennessee", x = "Event Type", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for better readability

#total = 82

```

```{r}
#how much damage was done
dam_tenn <- noaa_20_21 |>
  filter(STATE == "TENNESSEE") |>
  summarise(DAMAGE_TOTAL = sum(DAMAGE_TOTAL))
#Answer = $521,911,000

#no need to graph. just curious
  

#####INVESTIGATE

```

#damage vs death
```{r}

interact <- ggplot(noaa_20_21, aes(x = DAMAGE_TOTAL, y = DEATH_TOTAL, text = STATE, fill = STATE)) +
  geom_point() +
  labs(title = "Flood Damage vs. Death Total",
       x = "Flood Damage",
       y = "Death Total"
       ) +
  theme_minimal()

ggplotly(interact)

#There is no correlation between amount of damage and deaths
```

#sometime between 2020-2021 there was a flood that caused 19 total deaths and $10,000,000 worth of damage AUGUST 21/ 2021 HUMPHREYS COUNTY

extra questions
-what qualifies as damage

-what causes a lakeshore flood

-what causes a coastal flood

-why does washington and new hamphire have no flood damage


-what states have what types of flood causes

Future work
- graph on a US map to get a better idea
- demographics of regions and how that plays a role on relief given
- flood insurance need vs. have
- how does was Wisconsin affected by a tropical system?
```{r}

state_trop <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Heavy Rain / Tropical System") |>
  group_by(STATE)|>
  select(STATE, YEAR, MONTH_NAME, FLOOD_CAUSE)|>
  distinct(STATE, .keep_all = TRUE)
#would assume states along the southern and eastern coast. How did Wisconsin get in here?


kable(state_trop, align = "c", caption = "TROPICAL SYSTEM STATES") #create table

state_ice <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Ice Jam")|>
  group_by(STATE)
#makes sense. Northern colder states

state_burn <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Heavy Rain / Burn Area")|>
  group_by(STATE)
#makes sense. Hot South Western states

state_melt <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Heavy Rain / Snow Melt")|>
  group_by(STATE)
#makes sense. Colder Northern states

state_levee <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Dam / Levee Break")|>
  group_by(STATE)
#makes sense. Any state with a Dam / Levee

state_release <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Planned Dam Release")|>
  group_by(STATE)
#makes sense. Any state with a Dam

state_just_rain <- noaa_20_21 |>
  filter(FLOOD_CAUSE == "Heavy Rain")|>
  group_by(STATE)
#makes sense


#I want to visualize this make a table?
```

###<FEMA>###

#load in the FEMA data and merge
```{r}
Mydata_3 <- read_csv("FemaWebDisasterSummaries.csv")
Mydata_4 <- read_csv("DisasterDeclarationsSummaries.csv")
fema_data <- left_join(Mydata_3, Mydata_4, "disasterNumber")
#how much relief was sent
```

#filter to keep only columns of interest
```{r}

fema_data_int <- fema_data |> 
  filter(incidentType == "Flood") |>
  filter(incidentBeginDate == "2020" | incidentBeginDate ==  "2021")|>
  select(-hash.x, -lastRefresh.x, -id.x, -declarationType, -disasterCloseoutDate, 
         -tribalRequest, -lastRefresh.y, -id.y, -hash.y, -iaLoadDate, -lastIAFilingDate, -ihProgramDeclared, -iaProgramDeclared, -paProgramDeclared, -hmProgramDeclared)|>
  relocate(designatedArea, .before = declarationDate)


```

```{r}
#how much relief was given to one the deadliest floods
tenn_relief <- fema_data_int |>
  filter(state == "TN")

```


